### 2018-01-17 manuscript.rmd
added Bioproject accession
light edits throughout, did not rebuild pdf

Siphonophores are a diverse group of hydrozoans (Cnidaria) that are found at all depths of the ocean - from the surface of the ocean, like the familiar Portuguese man of war, to the deep sea. Siphonophores play an important role in ocean ecosystems, and are among the most abundant gelatinous predators in the open ocean. Previous work with a few molecular markers provided insight into the internal relationships between major siphonophore groups, however there was little support for many deep relationships within the clade Codonophora. Here, we present a new siphonophore phylogeny, obtained by collecting new transcriptome data from 30 siphonophore species, and analyzing them in combination with 13 publicly available genomic and transcriptomic datasets. We use this new phylogeny to reconstruct several traits that are central to siphonophore biology, including gain and loss of zooid types, life history traits, and habitat. Sequence assembly, annotation, homology evaluation, and supermatrix construction were all conducted using the Agalma pipeline, and the supermatrix was subsequently analysed using Maximum Likelihood and Bayesian approaches. The phylogenetic relationships in this study are largely consistent with the previous phylogeny, but finds strong support for new clades within Codonophora that were previously unresolved. These results have important implications for trait evolution within Siphonophora, including favoring the hypothesis that monoecy arose twice.

@@

Siphonophores (Fig. 1 and 2) are among the most abundant gelatinous predators in the open ocean, and have a large impact on ocean ecosystems [@pugh1984diel;@pugh1997vertical; @williams1981vertical; @purcell1981dietary; @pages2001gelatinous;  @Choy20172116]. Siphonophores, which belong to Hydrozoa (Cnidaria), are found at all depths in the ocean. The most familiar species is the Portuguese man of war *Physalia physalis*, which floats at the surface and can wash up conspicuously onto beaches [@Totton1960]. Most species are planktonic, living in the water column, where some grow to be more than 30 meters in length [@mackie1987siphonophore]. There is also a small clade of benthic siphonophores, the Rhodaliidae [@pugh1983benthic], that are tethered to the bottom for part of their lives. There are currently 188 valid described siphonophore species.

@@

Understanding the unique ecology, morphology, and development of siphonophores requires a well-resolved phylogeny of the group. The relationship of siphonophores to other hydrozoans has been difficult to resolve [@cartwright2010character; @Cartwright2008kl; @zapata2015phylogenomic], but there has been progress on their internal relationships. A phylogeny [@Dunn:2005dy] based on two genes (16S, 18S) from 52 siphonophore taxa advanced several long standing questions about siphonophore biology. These include the relationships of the three historically recognised groups, Cystonectae, Physonectae, and Calycophorae. The Cystonectae was found to be sister to all other siphonophores, while the Calycophorae were nested within "Physonectae". The name Codonophora was given to this clade of "Physonectae" and "Calycophorae" [@Dunn:2005dy].

@@

Most relationships received strong support across analysis methods (Fig. 3A), with a couple of localized exceptions (Fig. 3B). The phylogenetic relationships recovered in this study are largely consistent with those found in a previous study based on two genes (16S and 18S ribosomal RNA) [@Dunn:2005dy]. Relationships that receive strong support in both studies include the placement of Cystonectae as sister to Codonophora (the clade that includes all other siphonophores), the placement of Apolemiidae as sister to all other codonophorans, and the placement of Calycophorae within the paraphyletic "Physonectae". Multiple nodes that were not resolved in the previous two-gene analysis receive strong support in the present 1,071-gene transcriptome analyses. There is strong support for Pyrostephidae as sister to all other non-apolemiid codonophorans. Within the clade that is sister to the Pyrostephidae, we find two main clades, Calycophorae and a clade we here name Euphysonectae (Fig. 3A). It includes the remaining non-apolemiid, non-pyrostephid "Physonectae". We define Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to *Diphyes dispar*.

Euphysonectae consists of two reciprocally monophyletic groups that we here provisionally refer to as Clade A and Clade B (Fig. 3A). Clade A is a group of siphonophores that all have an involucrum, or fold, that forms around the base of the cnidoband [@totton1965synopsis]. This clade also shares the prescence of a descending mantle canal within the nectophores (Fig. S6), and is also monoecious (Fig. 5). Within Clade B there is low support for the placement of *Erenna richardi*, which is placed as sister to Clade A in some analyses (Fig. 3B). More taxon sampling will be required to determine the relationship of species within this clade.

Within Clade A, *Physophora gilmeri* along with *Lychnagalma utricularia* (both not included in the previous phylogeny) are sister to the Agalmatidae, a clade restricted to *Agalma*, *Athorybia*, *Melophysa*, *Halistemma* and *Nanomia* [@Dunn:2005dy; @pugh2006taxonomic]. In the rDNA study, *P. hydrostatica* (the presumed sister species to *P. gilmeri*) was sister to the Forskaliidae with low support. The position of *Cordagalma cordiforme* (= *Cordagalma ordinatum*) [@pugh2016synopsis] was previously unresolved, while in this analysis *Cordagalma* sp. is in a clade with *Forskalia asymmetrica*, falling outside of the Agalmatidae. Placement of *Cordagalma* outside Agalmatidae is consistent with previous analyses of morphological and molecular data [@pugh2006taxonomic; @Dunn:2005dy].

@@

![(A) Maximum likelihood (ML) phylogram with bipartation frequencies from the ML bootstraps and the Bayesian posterior distribution of trees. Unlabeled nodes have support >0.99 for both bootstraps and posteriors. (B) The topologies found in the psoterior distribution of trees that conflict with the ML tree. (C) The topologies evaluated by the SOWH tests. For more details on the SOWH topologies refer to Fig. S2.](figures/Figure3.pdf)

Using the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic], we evaluated the following three alternative phylogenetic hypotheses against the most likely tree topology (Fig. 3C): (i) monophyletic Physonectae, (ii) and (iii) monophyletic monoecious siphonophores,with and without *Rudjakovia* sp. respectively. In all three tests the alternative hypothesis was rejected (p-value <0.01, confidence interval: <0.001 - 0.03, Fig. S2).

The broad sampling approach of this phylogeny provides new evidence for the relationships between major siphonophore clades within Codonophora, specifically between Pyrostephidae, Calycophorae, and the newly named Euphysonectae. This opens up new questions about key relationships within both Calycophorae and Euphysonectae -- where future transcriptome sampling efforts should be focussed. Within Euphysonectae, two clades (Clade A and Clade B) are hypothesized, however there is weaker support for Clade B (Fig. 3A, 3B). Expanding sampling of species that probably fall in Clade B, including other *Erenna* species, rhodaliids, and relatives of Undescribed sp L, will greatly expand our understanding of these two groups and perhaps provide evidence of Clade B synapomorphies. Similarly, within Calycophorae, increased taxon sampling is needed. This study, and the previous study, suggest that the prayomorphs are paraphyletic, but for slightly different reasons. In @Dunn:2005dy, a clade of prayomorphs including *Praya dubia*, *Nectadamas diomedeae*, and *Nectopyramis natans* (not included in this study) were found to be sister to all other calycophorans, while in this study, the prayomorph *Lilyopsis fluoracantha* (not included in the previous study) is found in a clade including diphyomorph calycophorans that is sister to all other prayomorphs. Expanded taxon sampling, particularly *P. dubia* or a nectopyramid, but also extensive sampling across the major prayomorph and diphyomorph groups, will expand our understanding of relationships within Calycophorae. This will be especially important for understanding trait evolution within Calycophorae,  for example, the release of eudoxids (Fig. 4), or the arrangement of male and female zooids along the stem (see section 3.2 below).

@@

In all siphonophores, each gonophore (sexual medusa that produces gametes) is either male or female. Within each siphonophore species, colonies are either monoecious (male and female gonophores are on the same colony) or dioecious (male and female gonophores are on different colonies). Previous analyses suggested that the common ancestor of siphonophores was dioecious, and was consistent with a single gain of monoecy within the Codonophora and no secondary losses [@Dunn:2005dy]. The better-resolved tree in the current analyses indicates that the evolution of monoecy is more complicated than this. The two clades of monophyloetic siphonophores, Calycophorae and Clade A (Fig. 3A), do not form a monophyletic group. This is because Clade B, which contains dioecious species, is also desceneded from their most recent common ancestor. SOWH tests strongly reject the placement of the monecious clades Calycophorae and Clade A as a group that excludes Clade B (Figs. 3C and S2). The positions of the only two taxa from Clade B in the previous analysies [@Dunn:2005dy], *Erenna* and *Stephalia*, were unresolved in that previous study. This difference in conclusions regarding trait evolution, therefore, does not reflect a contradication between alternative strongly supported results, but the resolution of earlier polytomies in a way that indicates there has been homoplasy in the evolution of monoecy.

@@

Within the nectosome, the nectophores can be attached along the dorsal or ventral side of the stem, following the orientation framework of @haddock_dunn_pugh_2005. Our ancestral reconstructions for this character (Fig. S7) suggest that ventral attachment of nectophores was the ancestral state in Codonophora, and that dorsal attachment has independently evolved twice -- once along the Agalmatidae stem and once along the Pyrostephidae stem.

@@

Siphonophores are abundant predators in the pelagic realm, ranging from the surface (*Physalia physalis*) to bathypelagic depths (Figs. 4 and S8) [@mapstone2014global; @mackie1987siphonophore]. The depth distribution of siphonophore populations is not always static, as some species are known to be vertical migrators, although this is within a relatively narrow depth range (<100m) [@pugh1984diel]. Some species such as *Nanomia bijuga* exhibit synchronous diel migration patterns [@barham1966deep]. Using the present phylogeny, we reconstructed the median depth changes along the phylogeny under a Brownian Motion model (Fig. S9), which had the strongest AICc support (compared to non-phylogenetic distributions, and to Ohrnstein-Uhlenbeck). This model indicates a mesopelagic most recent common ancestor, with several independent transition events to epipelagic (light blue) and bathypelagic (dark blue) waters. There was only a single transition to benthic lifestyle on the branch of Rhodaliidae, and a single transition to a pleustonic lifestyle on the branch of *Physalia physalis*. Depth appears to be phylogenetically conserved in Euphysonectae after the split between Clade A (shallow living species) and Clade B (deep dwelling species); however several shallow-living species that likely belong in Clade B were not included in this analysis, suggesting that depth may be more labile than it appears. Phylogenetic conservatism is also observed for the mesopelagic Apolemiidae and Pyrostephidae clades, but not for the Calycophorae, which show a diversified evolutionary trajectory for depth. The present sampling, however, is not sufficient to capture significant variation in depth distributions between closely related species. Previous studies have shown that many species that are collected at the same locality are found to occupy discrete, largely non-overlapping depth distributions, including between species that are closely related [@pugh1974vertical].

>>>>

Siphonophores are a diverse group of hydrozoans (Cnidaria) that are found at all depths of the ocean - from the surface of the ocean, like the familiar Portuguese man of war, to the deep sea. Siphonophores play an important role in ocean ecosystems, and are among the most abundant gelatinous predators in the open ocean. A previous phylogenetic study based on two ribosomal RNA genes provided insight into the internal relationships between major siphonophore groups, however there was little support for many deep relationships within the clade Codonophora. Here, we present a new siphonophore phylogeny based on new transcriptome data from 30 siphonophore species analyzed in combination with 13 publicly available genomic and transcriptomic datasets. We use this new phylogeny to reconstruct several traits that are central to siphonophore biology, including sexual system (monoecy vs. dioecy), gain and loss of zooid types, life history traits, and habitat. The phylogenetic relationships in this study are largely consistent with the previous phylogeny, but finds strong support for new clades within Codonophora that were previously unresolved. These results have important implications for trait evolution within Siphonophora, including favoring the hypothesis that monoecy arose twice.

@@

Siphonophores (Fig. 1 and 2) are among the most abundant gelatinous predators in the open ocean, and have a large impact on ocean ecosystems [@pugh1984diel;@pugh1997vertical; @williams1981vertical; @purcell1981dietary; @pages2001gelatinous;  @Choy20172116]. Siphonophores, which belong to Hydrozoa (Cnidaria), are found at all depths in the ocean. The most familiar species is the Portuguese man of war *Physalia physalis*, which floats at the surface and can wash up conspicuously onto beaches [@Totton1960]. Most species are planktonic, living in the water column, where some grow to be more than 30 meters in length [@mackie1987siphonophore]. There is also a small clade of benthic siphonophores, Rhodaliidae [@pugh1983benthic], that are tethered to the bottom for part of their lives. There are currently 188 valid described siphonophore species.

@@

Understanding the unique ecology, morphology, and development of siphonophores requires a well-resolved phylogeny of the group. The relationship of siphonophores to other hydrozoans has been difficult to resolve [@cartwright2010character; @Cartwright2008kl; @zapata2015phylogenomic], but there has been progress on their internal relationships. A phylogeny [@Dunn:2005dy] based on two genes (16S, 18S) from 52 siphonophore taxa advanced several long standing questions about siphonophore biology. These include the relationships of the three historically recognised groups, Cystonectae, Physonectae, and Calycophorae. Cystonectae was found to be sister to all other siphonophores, while Calycophorae were nested within "Physonectae". The name Codonophora was given to this clade of "Physonectae" and "Calycophorae" [@Dunn:2005dy].

@@

Most relationships received strong support across analysis methods (Fig. 3A), with a couple of localized exceptions (Fig. 3B). The phylogenetic relationships recovered in this study are largely consistent with those found in a previous study based on two genes (16S and 18S ribosomal RNA) [@Dunn:2005dy]. Relationships that receive strong support in both studies include the placement of Cystonectae as sister to Codonophora (the clade that includes all other siphonophores), the placement of Apolemiidae as sister to all other codonophorans, and the placement of Calycophorae within the paraphyletic "Physonectae". Multiple nodes that were not resolved in the previous two-gene analysis receive strong support in the present 1,071-gene transcriptome analyses. There is strong support for Pyrostephidae as sister to all other non-apolemiid codonophorans. Within the clade that is sister to Pyrostephidae, we find two main clades, Calycophorae and a clade we here name Euphysonectae (Fig. 3A). It includes the remaining non-apolemiid, non-pyrostephid "Physonectae". We define Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to *Diphyes dispar*.

Euphysonectae consists of two reciprocally monophyletic groups that we here provisionally refer to as Clade A and Clade B (Fig. 3A). The presence of an involucrum, a fold around the base of the cnidoband [@totton1965synopsis], is a potential synapomorphy for Clade A. Species of Clade A also have a descending mantle canal within the nectophores (Fig. S6), a structure that is also present in some calycophorans. Members of Clade A are also monoecious (Fig. 5). There is not a clear synapomorphy for Clade B. Within Clade B there is low support for the placement of *Erenna richardi*, which is placed as sister to Clade A in some analyses (Fig. 3B). More taxon sampling will be required to determine the relationship of species within this clade.

Within Clade A, *Physophora gilmeri* along with *Lychnagalma utricularia* (both not included in the previous phylogeny) are sister to Agalmatidae, a clade restricted to *Agalma*, *Athorybia*, *Melophysa*, *Halistemma* and *Nanomia* [@Dunn:2005dy; @pugh2006taxonomic]. In the rDNA study, *P. hydrostatica* (the presumed sister species to *P. gilmeri*) was sister to Forskaliidae with low support. The position of *Cordagalma cordiforme* (= *Cordagalma ordinatum*) [@pugh2016synopsis] was previously unresolved, while in this analysis *Cordagalma* sp. is in a clade with *Forskalia asymmetrica*, falling outside of Agalmatidae. Placement of *Cordagalma* outside Agalmatidae is consistent with previous analyses of morphological and molecular data [@pugh2006taxonomic; @Dunn:2005dy].

@@

![(A) Maximum likelihood (ML) phylogram with bipartation frequencies from the ML bootstraps and the Bayesian posterior distribution of trees. Unlabeled nodes have support >0.99 for both bootstraps and posteriors. (B) The topologies found in the posterior distribution of trees that conflict with the ML tree. (C) The topologies evaluated by the SOWH tests. For more details on the SOWH topologies refer to Fig. S2.](figures/Figure3.pdf)

Using the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic], we evaluated the following three alternative phylogenetic hypotheses against the most likely tree topology (Fig. 3C): (i) monophyletic Physonectae, (ii) and (iii) monophyletic monoecious siphonophores, with and without *Rudjakovia* sp. respectively. In all three tests the alternative hypothesis was rejected (p-value <0.01, confidence interval: <0.001 - 0.03, Fig. S2).

The broad sampling approach of this phylogeny provides new evidence for the relationships between major siphonophore clades within Codonophora, specifically between Pyrostephidae, Calycophorae, and the newly named Euphysonectae. This opens up new questions about key relationships within both Calycophorae and Euphysonectae -- where future transcriptome sampling efforts should be focussed. Within Euphysonectae, two clades (Clade A and Clade B) are hypothesized, however there is weaker support for Clade B (Fig. 3A, 3B). Expanding sampling of species that probably fall in Clade B, including other *Erenna* species, rhodaliids, and relatives of Undescribed sp L, will greatly expand our understanding of these two groups and perhaps provide evidence of Clade B synapomorphies. Similarly, within Calycophorae, increased taxon sampling is needed. This study, and the previous phylogenetic study [@Dunn:2005dy], suggest that the prayomorphs are paraphyletic, but for slightly different reasons given the different sampling of the analyses. In @Dunn:2005dy, a clade of prayomorphs including *Praya dubia*, *Nectadamas diomedeae*, and *Nectopyramis natans* (not included in this study) were found to be sister to all other calycophorans, while in this study, the prayomorph *Lilyopsis fluoracantha* (not included in the previous study) is found in a clade including diphyomorph calycophorans that is sister to all other prayomorphs. Expanded taxon sampling, particularly *P. dubia* or a nectopyramid, but also extensive sampling across the major prayomorph and diphyomorph groups, will expand our understanding of relationships within Calycophorae. This will be especially important for understanding trait evolution within Calycophorae,  for example, the release of eudoxids (Fig. 4), or the arrangement of male and female zooids along the stem (see section 3.2 below).

@@

In all siphonophores, each gonophore (sexual medusa that produces gametes) is either male or female. Within each siphonophore species, colonies are either monoecious (male and female gonophores are on the same colony) or dioecious (male and female gonophores are on different colonies). Previous analyses suggested that the common ancestor of siphonophores was dioecious, and was consistent with a single gain of monoecy within Codonophora and no secondary losses [@Dunn:2005dy]. The better-resolved tree in the current analyses indicates that the evolution of monoecy is more complicated than this. The two clades of monophyloetic siphonophores, Calycophorae and Clade A (Fig. 3A), do not form a monophyletic group. This is because Clade B, which contains dioecious species, is also desceneded from their most recent common ancestor. SOWH tests strongly reject the placement of the monecious clades Calycophorae and Clade A as a group that excludes Clade B (Figs. 3C and S2). The positions of the only two taxa from Clade B in the previous analysies [@Dunn:2005dy], *Erenna* and *Stephalia*, were unresolved in that previous study. This difference in conclusions regarding trait evolution, therefore, does not reflect a contradication between alternative strongly supported results, but the resolution of earlier polytomies in a way that indicates there has been homoplasy in the evolution of monoecy.

@@

Within the nectosome, the nectophores can be attached along the dorsal or ventral side of the stem, following the orientation framework of @haddock_dunn_pugh_2005. Our ancestral reconstructions for this character (Fig. S7) suggest that ventral attachment of nectophores was the ancestral state in Codonophora, and that dorsal attachment has independently evolved twice -- once along the stem of Agalmatidae and once along the stem of Pyrostephidae.

@@

Siphonophores are abundant predators in the pelagic realm, ranging from the surface (*Physalia physalis*) to bathypelagic depths (Figs. 4 and S8) [@mapstone2014global; @mackie1987siphonophore]. The depth distribution of siphonophore populations is not always static, as some species are known to be vertical migrators, although this is within a relatively narrow depth range (<100m) [@pugh1984diel]. Some species such as *Nanomia bijuga* exhibit synchronous diel migration patterns [@barham1966deep]. Using the present phylogeny, we reconstructed the median depth changes along the phylogeny under a Brownian Motion model (Fig. S9), which had the strongest AICc support (compared to non-phylogenetic distributions, and to Ohrnstein-Uhlenbeck). This model indicates a mesopelagic most recent common ancestor, with several independent transition events to epipelagic (light blue) and bathypelagic (dark blue) waters. There was only a single transition to benthic lifestyle on the branch of Rhodaliidae, and a single transition to a pleustonic lifestyle on the branch of *Physalia physalis*. Depth appears to be phylogenetically conserved in Euphysonectae after the split between Clade A (shallow living species) and Clade B (deep dwelling species); however several shallow-living species that likely belong in Clade B were not included in this analysis, suggesting that depth may be more labile than it appears. Phylogenetic conservatism is also observed for the mesopelagic Apolemiidae and Pyrostephidae clades, but not for Calycophorae, which show a diversified evolutionary trajectory for depth. The present sampling, however, is not sufficient to capture significant variation in depth distributions between closely related species. Previous studies have shown that many species that are collected at the same locality are found to occupy discrete, largely non-overlapping depth distributions, including between species that are closely related [@pugh1974vertical].
